---
title: "Transcriptomics and the analysis of RNA-Seq data"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import today's data

The data for this hands-on session comes from a published RNA-seq experiment where airway smooth muscle cells were treated with dexamethasone, a synthetic glucocorticoid steroid with anti-inflammatory effects (Himes et al. 2014).

Glucocorticoids are used, for example, by people with asthma to reduce inflammation of the airways. The anti-inflammatory effects on airway smooth muscle (ASM) cells has been known for some time but the underlying molecular mechanisms are unclear.

```{r}
counts <- read.csv("airway_scaledcounts (1).csv", stringsAsFactors = FALSE)
metadata <-  read.csv("airway_metadata.csv", stringsAsFactors = FALSE)
```

How many genes are we dealing with?

```{r}
nrow(counts)
```

We want to calculate the average counts per gene for contorl samples and the for the drug treated samples. We will then compare these values to see if there is a difference overall between control and treated samples for a given gene.

This is our `metadata` info that tells mw where our control and treated samples are (i.e. which columns in `counts`)
```{r}
metadata
```

Now I want to access the control only rows in the metadata 
```{r}
control <- metadata[metadata$dex == "control", ]
control
#metadata$id[ metadata$dex == "control" ]
```

Now use the $id of the control data to access the control columns in `counts and calculate the average.
```{r}
control.mean <- rowSums( counts[ , control$id] ) / nrow(control)
names(control.mean) <- counts$ensgene
```

Now lets do the same for *treated* samples
```{r}
treated <- metadata[metadata$dex == "treated",]
treated.mean <- rowSums(counts[,treated$id]) / nrow(treated)
```

We will combine our meancount data for bookkeeping purposes.
```{r}
meancounts <- data.frame(control.mean, treated.mean)
head( meancounts )
```

Lets plot control.mean vs treated.mean

```{r}
plot(meancounts)
```

Or with ggplot

```{r}
library(ggplot2)
ggplot(meancounts, aes(x=control.mean, y=treated.mean)) +
  geom_point()
```

```{r}
ggplot(meancounts, aes(control.mean)) + 
  geom_density()
```

Lets make this plot a log-log plot
```{r}
plot(meancounts, log="xy")
```

```{r}
ggplot(meancounts, aes(x=control.mean, y=treated.mean)) +
  geom_point() + 
  scale_x_log10() + 
  scale_y_log10() + 
  annotation_logticks()
```

```{r}
ggplot(meancounts, aes(control.mean)) + 
  geom_density() + 
  scale_x_log10() +
  annotation_logticks()
```

```{r}
ggplot(meancounts) + 
  geom_density(aes(control.mean, color="red")) + 
  geom_density(aes(treated.mean), color="blue") +
  scale_x_log10() +
  annotation_logticks()
```

We can find candidate differentially expressed genes by looking for genes with a large change between control and dex-treated samples. We usually look at the log2 of the fold change, because this has better mathematical properties.

Here we calculate log2foldchange, add it to our meancounts data.frame

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

Remove any gene that has zero average counts.

```{r}
x <- data.frame(c(1,1,10,0), c(0,1,10,0))
#x
#x==0
ind <-unique( which(x == 0, arr.ind=TRUE)[,"row"] )
x[-ind,]
```



```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

How many genes do I have left?

```{r}
nrow(mycounts)
```

A common threshold used for calling something differentially expressed is a log2(FoldChange) of greater than 2 or less than -2. Let’s filter the dataset both ways to see how many genes are up or down-regulated.

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

```{r}
sum(up.ind)
sum(down.ind)
```

## Do this properly with DESeq2

```{r}
library(DESeq2)

```

Setup input for DESeq by specifing countdata, metadata and the experiment design (i.e. which comparasions we are intrested in)

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex, 
                              tidy=TRUE)
dds
```

```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
```

```{r}
plot(res$log2FoldChange, res$padj)
```

```{r}
plot(res$log2FoldChange, log(res$padj))
```

Lets plot this the way the rest of the world does!
```{r}
plot(res$log2FoldChange, -log(res$padj))
```


```{r}
mycols <- rep("gray", nrow(res))
mycols[ res$log2FoldChange > 2 ] <- "blue"
mycols[ res$log2FoldChange < -2 ] <- "blue"

plot(res$log2FoldChange, -log(res$padj), col=mycols)
abline(v=c(2,-2), lty=2)
abline(h=-log(0.05) )
```







